---
title: "BiocExecute"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BiocExecute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BiocExecute)
```

# Motivation

FAIR

+ for devs

+ for users

Use cases

# How to create executables

`BiocExecute` uses `Rapp` to make your package functions executable. In the
coming sections we will show the different ways `Rapp` includes arguments to be
called in the CLI. Note that `Rapp` by itself works with scripts in which the
first line is defined as `#!/usr/bin/env Rapp`. You should **_NOT_** include
this line in your scripts ! This is handled internally as it is bundled in the
`BiocExecute` package.



## Create a script

### Exec

### Describe Rapp fields (description, parameters)


## Create a install function

-> Using BiocExecute::install_cli that uses Rapp:::install_pkg_cli_apps

## Explain the installation process to the user


# 
