---
title: "BiocExecute"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BiocExecute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BiocExecute)
```

# Commandline executables for bioconductor functions and scripts

This project was initiated during the [EuroBioC26 Hackathon](https://github.com/BiocCodingCollaborations/EuroBioc2026_Hackathon).

[Vision](https://github.com/BiocCodingCollaborations/EuroBioc2026_Hackathon#commandline-executables-for-bioconductor-functions-and-scripts)

# Motivation

* Bioconductor packages

  * > 2,400 packages (release 3.23)
  * Million downloads per year
  * Thoroughly documented and maintained

* Can be even better --> FAIRification

  * Findability: 
  * Accessibility: command line makes it usable by more users in more contexts
  * Interoperability: combine with other command line tools not necessarily R tools
  * Reusability: modular components to be used/reused in new workflows (nextflow, snakemake, whatever) like independent building bricks

* Use cases
  * User: how can I use this cool package in my workflow (nextflow or else)
  * Developer: how can my package be used and downloaded by more users

* Benefits / new relevant features

  * Light weight, relies on existing packages 
  * Expand usability of packages


# How to create executables

`BiocExecute` uses `Rapp` to make your package functions executable. In the
coming sections we will show the different ways `Rapp` includes arguments to be
called in the CLI. Note that `Rapp` by itself works with scripts in which the
first line is defined as `#!/usr/bin/env Rapp`. You should **_NOT_** include
this line in your scripts ! This is handled internally as it is bundled in the
`BiocExecute` package.



## Create a script

### Exec

### Describe Rapp fields (description, parameters)


## Create a install function

-> Using BiocExecute::install_cli that uses Rapp:::install_pkg_cli_apps

## Explain the installation process to the user


# 
