---
title: "BiocExecute: a package to use upon package building for creating executables."
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{BiocExecute}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{BiocExecute}
    %\VignetteDepends{Rapp}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("BiocExecute")`<br />
**Authors**: `r packageDescription("BiocExecute")[["Author"]] `<br />
**Last modified:** `r file.info("BiocExecute.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("BiocExecute")
```

# Commandline executables for bioconductor functions and scripts

This project was initiated during the [EuroBioC26
Hackathon](https://github.com/BiocCodingCollaborations/EuroBioc2026_Hackathon).

[Vision](https://github.com/BiocCodingCollaborations/EuroBioc2026_Hackathon#commandline-executables-for-bioconductor-functions-and-scripts)

# Motivation

Bioconductor has a collection of more than 2,400 open source software packages
downloaded millions of times per year. They are thoroughly maintained and
documented, and their quality is ensured through the use of BiocCheck.

This package aims to further improve Bioconductor software FAIRness, by making
them usable in the command line.

* Findability: xxx;

* Accessibility: allows more users to use Bioconductor packages, in a wider
variety of setups;

* Interoperability: packages can be combined as modules with other command line
tools;

* Reusability: modular components can be reused in future workflows with any
workflow manager.

Package users and developers can both benefit from this setup: users can use
Bioconductor tools out side of R scripts and integrate them in their own
workflows, and package developers can benefit from a wider range of users.
Overall, Bioconductor software can gain visibility among a larger community of
bioinformaticiens.

Light weight, relies on existing packages 

# How to use executables

Here's a quick example of how you would call the function `name(x, y)` from
the package `pkgExample`:

First, the user needs to make sure the package functions are executable:

```{r inst, eval = FALSE}
BiocExecute::installExecs("pkgExample")
```

And now call those functions from within the terminal:

```{bash ex, eval = FALSE}
pkgExample --help
pkgExample name -x Bilbo -y Baggins
```

# How to create executables

`BiocExecute` uses `Rapp` to make your package functions executable. In the
coming sections we will show the different ways `Rapp` includes arguments to be
called in the CLI. Note that `Rapp` by itself works with scripts in which the
first line is defined as `#!/usr/bin/env Rapp`. You should **_NOT_** include
this line in your scripts ! This is handled internally as it is bundled in the
`BiocExecute` package.

The executables can have many ranges, from a simple function call to an entire
complex workflow. Note that bigger workflows mean more parameters to call in
the CLI. 

As a maintainer, you may choose which functions/workflows you decide to include
in your package. As a package user, we suggest to create pull requests on the
package GitHub repository for workflows that you believe should be easily 
accessible. Try and avoid creating strongly personal workflows. Hold priority
for workflows that are repeatedly used across different user cases.

## Create a script

An _R_ package that has executables should include them in its `exec/scripts/`
directory. The maintainer of a package should then call
`BiocExecute::compileExecs()` to create the file `packageName.R` in the `exec/`
directory.

For instance, the name of my package is `mypkg`. In its root directory, I don't
have an `exec` directory:

```{r tree, eval = FALSE}
mypkg
│   README.md
│   DESCRIPTION    
│   NEWS.md
│   NAMESPACE    
│
└───R
│   │   functionA.R
│   │   functionB.R
│   
└───tests
│   │   testA.R
│   │   testB.R
│   
└───vignettes
    │   myVignette.Rmd
```

To create the necessary files, I use:

```{r createSkeleton, eval = FALSE}
## From the root directory
exec.skeleton("./")
```

Now, my directory tree looks like this:

```{r tree2, eval = FALSE}
mypkg
│   README.md
│   DESCRIPTION    
│   NEWS.md
│   NAMESPACE    
│
└───R
│   │   functionA.R
│   │   functionB.R
│   
└───exec
│   │   mypkg.R
│   │
│   └───scripts
│       │   template.R
│   
└───tests
│   │   testA.R
│   │   testB.R
│   
└───vignettes
    │   myVignette.Rmd
```

In the `exec` directory, there is a file called `mypkg.R`. This file is built
by compiling all scripts in `exec/scripts/`. It is this file that is
executable and should not be edited by hand as it will be overwritten by
`combineExecs()`.

### Scripts files

The files in the `exec/scripts/` directory are at the core of the available
executables. One script correcsponds to one command, which can be a simple
function or even a whole workflow. These scripts are combined and compiled into
the main executable file in `exec/` (see section above). 

The script files need to follow the `Rapp` architecture. A template file is
built by default to get you started. In practice, these _R_ scripts are really
a repetition of the important functions within your package, except that their
parameters and documentation are re-written in a way for `Rapp` to parse them.

### Rapp fields and function parameters

`Rapp` parses _R_ scripts by looking for specific expression patterns at the
top level. Each pattern maps to a different CLI surface. The table below
summarises the most common ones:

| R expression | CLI surface |
|---|---|
| `foo <- ""` | Option: `app --foo value` |
| `foo <- NULL` | Positional argument: `app foo-value` |
| `foo <- TRUE` | Boolean switch: `app --foo` / `app --no-foo` |
| `foo <- c()` | Repeatable option (raw strings): `app --foo a --foo b` |
| `foo <- list()` | Repeatable option (parsed values): `app --foo 1 --foo 2` |
| `switch("", cmd1 = {}, cmd2 = {})` | Subcommands: `app cmd1 --help` |

Annotations are written as YAML hash-pipe comments (`#|`) directly above the
assignment they document. The most commonly used fields are:

| Field | Description |
|---|---|
| `description` | Short description shown in `--help` output |
| `title` | Title for subcommands |
| `short` | Single-letter alias (e.g. `short: n` enables `-n`) |
| `required` | Set to `false` to make a positional argument optional |
| `val_type` | Expected type: `string`, `integer`, `float`, `bool`, or `any` |

A minimal example script illustrates how these come together:

```r
#!/usr/bin/env Rapp
#| description: Count word occurrences in a file.

#| description: Path to the input file.
inputFile <- NULL

#| description: Word to count.
#| short: w
word <- ""

#| description: Print each match.
verbose <- FALSE
```

Called from the terminal:

```bash
count-words myfile.txt --word hello --verbose
count-words --help
```

For a full description of all available fields and advanced patterns such as
nested subcommands, refer to the [`Rapp` GitHub
page](https://github.com/r-lib/Rapp).
