---
title: "BiocExecute: a package to use upon package building for creating executables."
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{BiocExecute}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{BiocExecute}
    %\VignetteDepends{Rapp}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("BiocExecute")`<br />
**Authors**: `r packageDescription("BiocExecute")[["Author"]] `<br />
**Last modified:** `r file.info("BiocExecute.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("BiocExecute")
```

# Commandline executables for bioconductor functions and scripts

This project was initiated during the [EuroBioC26
Hackathon](https://github.com/BiocCodingCollaborations/EuroBioc2026_Hackathon).

[GitHub repo](https://github.com/BiocCodingCollaborations/EuroBioc2026_Hackathon#commandline-executables-for-bioconductor-functions-and-scripts)

# Motivation

Bioconductor has a collection of more than 2,400 open source software packages
downloaded millions of times per year. They are thoroughly maintained and
documented, and their quality is ensured through the use of BiocCheck.

This package aims to further improve Bioconductor software FAIRness, by making
them usable in the command line.

* Findability: not applicable;

* Accessibility: allows more users to use Bioconductor packages, in a wider
variety of setups;

* Interoperability: packages can be combined as modules with other command line
tools;

* Reusability: modular components can be reused in future workflows with any
workflow manager.

Package users and developers can both benefit from this setup: users can use
Bioconductor tools outside of R scripts and integrate them in their own
workflows, and package developers can benefit from a wider range of users.
Overall, Bioconductor software can gain visibility among a larger community of
bioinformaticiens.

Lightweight, relies on existing packages (`Rapp`).

# How to use executables

Here's a quick example of how you would call the tool `name(x, y)` from
the package `pkgExample` as a package _user_:

First, the user needs to make sure the package executables are available:

```{r inst, eval = FALSE}
BiocExecute::execInstall("pkgExample")
```

And now call those functions from within the terminal:

```{bash ex, eval = FALSE}
pkgExample --help
pkgExample name -x Bilbo -y Baggins
```

# How to create executables as a maintainer

`BiocExecute` uses `Rapp` to make your package functions/scripts executable. In
the coming sections we will show the different ways `Rapp` includes arguments
to be called in the CLI. Note that `Rapp` by itself works with scripts in which
the first line is defined as `#!/usr/bin/env Rapp`. You should **_NOT_**
include this line in your scripts ! This is handled internally as it is bundled
in the `BiocExecute` package.

The executables can have many ranges, from a simple function call to an entire
complex workflow. Note that bigger workflows mean more parameters to call in
the CLI. 

As a maintainer, you may choose which functions/workflows you decide to include
in your package. As a package user, we suggest to create pull requests on the
package GitHub repository for workflows that you believe should be easily 
accessible. Try and avoid creating strongly personal workflows. Hold priority
for workflows that are repeatedly used across different user cases.

## Create a script

An _R_ package that has executables should include them in its `exec/scripts/`
directory. `BiocExecute` has all the necessary functions to create these
folders, scripts and more.

For instance, suppose the name of my package is `mypkg`. In its root directory,
I don't have an `exec` nor a `scripts` folder:

```{r tree, eval = FALSE}
mypkg
│   README.md
│   DESCRIPTION    
│   NEWS.md
│   NAMESPACE    
│
└───R
│   │   functionA.R
│   │   functionB.R
│   
└───tests
│   │   testA.R
│   │   testB.R
│   
└───vignettes
    │   myVignette.Rmd
```

To create the necessary files, I use:

```{r createSkeleton, eval = FALSE}
## From the root directory
execSkeleton()
```

Now, my directory tree looks like this:

```{r tree2, eval = FALSE}
mypkg
│   README.md
│   DESCRIPTION    
│   NEWS.md
│   NAMESPACE    
│
└───R
│   │   functionA.R
│   │   functionB.R
│   
└───exec
│   │   mypkg.R
│   │
│   └───scripts
│       │   base_template.R
│   
└───tests
│   │   testA.R
│   │   testB.R
│   
└───vignettes
    │   myVignette.Rmd
```

For now, there is a simple template in the `scripts` folder. The same template
can be created using `execTemplate()`. 

The maintainer of a package should then create the scripts with functions or
workflows that he/she wishes to make executable on the CLI.

Once the `exec/scripts/` are created, you should (in order from the root
directory):

1. Call `execCompile()` to build the actual executable script. This file will
   be called by its package name and it will be written in the `exec` folder.
   Do not edit this file by hand, it is likely to be overwritten.
2. Call `devtools::install()` to re-install the package with the executables.
3. Call `execInstall("mypkg")` or a vector of packages to make the executables
   available in the CLI. To remove those, call `execUninstall("mypkg")`. To
   make the executables available system-wise (with sudo access), use the
   `destdir` parameter. 

Your package functions/tools are now available in the CLI !

### Scripts files

The files in the `exec/scripts/` directory are at the core of the available
executables. One script corresponds to one command, which can be a simple
function or even a whole workflow. These scripts are combined and compiled into
the main executable file in `exec/` (see section above). 

The script files need to follow the `Rapp` architecture. A template file is
built by default to get you started. In practice, these _R_ scripts are really
a repetition of the important functions within your package, except that their
parameters and documentation are re-written in a way for `Rapp` to parse them.

### Rapp fields and function parameters

`Rapp` parses _R_ scripts by looking for specific expression patterns at the
top level. Each pattern maps to a different CLI surface. The table below
summarises the most common ones:

| R expression | CLI surface |
|---|---|
| `foo <- ""` | Option: `app --foo value` |
| `foo <- NULL` | Positional argument: `app foo-value` |
| `foo <- TRUE` | Boolean switch: `app --foo=true` / `app --foo=true` |
| `foo <- c()` | Repeatable option (raw strings): `app --foo a --foo b` |
| `foo <- list()` | Repeatable option (parsed values): `app --foo 1 --foo 2` |
| `switch("", cmd1 = {}, cmd2 = {})` | Subcommands: `app cmd1 --help` |

Annotations are written as YAML hash-pipe comments (`#|`) directly above the
assignment they document. The most commonly used fields are:

| Field | Description |
|---|---|
| `description` | Short description shown in `--help` output |
| `title` | Title for subcommands |
| `short` | Single-letter alias (e.g. `short: n` enables `-n`) |
| `required` | Set to `false` to make a positional argument optional |
| `val_type` | Expected type: `string`, `integer`, `float`, `bool`, or `any` |

A minimal example script illustrates how these come together:

```r
#| description: Count word occurrences in a file.

#| description: Path to the input file.
inputFile <- NULL

#| description: Word to count.
#| short: w
word <- ""

#| description: Print each match.
verbose <- FALSE
```

Called from the terminal:

```bash
count-words myfile.txt --word hello --verbose
count-words --help
```

Use `execTemplate()` to create a template script with the different parameters
to facilitate writing your scripts.

For a full description of all available fields and advanced patterns such as
nested subcommands, refer to the [`Rapp` GitHub
page](https://github.com/r-lib/Rapp).
